### 2023-07-17 docs/manuscript.Rmd
improved manuscript

---
title             : "A Tutorial for Customized Power Simulation and Data Analysis using GLMM  - A Case Study for Human-AI Interaction Experiments"
shorttitle        : "GLMM Power Simulation Tutorial"

@@

affiliation:
  - id            : "1"
    institution   : "LMU Munich, Center for Leadership and People Management"
  - id            : "2"
    institution   : "LMU Munich, Department of Psychology"

@@

abstract: |
  Human-AI Intercation experiments are becoming increasingly important. However, many studies are underpowered since researchers lack the tools to conduct power analyses and sample size simulations. In this work, we provide. By providing code in a case study, we equip scientists to simulate their own data and run analyses based on generalized linear mixed models (GLMM).

keywords          : "generalized linear mixed model, data simulation, sample size, power analysis, data analysis"

@@

<!-- General intro, relevance of Human AI Interaction experiments -->

As the integration of artificial intelligence (AI) into our daily lives continues to increase, it has become increasingly important to study how humans interact with these intelligent systems. This is particularly important as AI becomes more sophisticated and is used in decision-making processes, for instance in medicine, that have significant impacts on individuals and society as a whole. Therefore, thourough experiments are needed to study human-AI interaction and understand how people perceive and respond to these technologies. When conducting research in this field, it is essential to determine an appropriate sample size to ensure that the results obtained are both statistically significant and reliable. Therefore, running sample size estimations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.

@@

<!-- Power simulations are hard! (especially for GLMMs) -->
Power simulations can be challenging, particularly when analyzing complex psychological study designs using (generalized) linear mixed models (GLMMs). While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of GLMMs in psychological research introduces greater complexity to power calculations. For GLMMs, simulations are often required, necessitating assumptions for both fixed and random effects. However, most tutorials on GLMMs primarily focus on the most common designs, limiting the guidance available for researchers facing more intricate scenarios. Moreover, existing tutorials often rely on heuristics for estimating random effect standard deviations and may incorporate meta-analytic results for both random and fixed effects, further adding to the challenge of conducting accurate power calculations for GLMMs.

@@

<!-- describe the lme4 package and how it works-->

The lme4 R package is a state-of-the-art tool for fitting frequentist Generalized Linear Mixed Models (GLMMs). It provides extensive capabilities for modeling complex data structures, including models with normally distributed random intercepts and random slopes. The package includes a useful function called "simulate" that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling power simulations and other related analyses.

@@

Human-AI interaction research requires meticulous planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and helper programs can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. In this paper, we discuss the necessity of bespoke power simulations in human-AI interaction research, their implications for research design, and the importance of teaching these skills to Human-AI interaction researchers. Furthermore, we highlight the discrepancy between the perceived effort involved in performing bespoke power simulations and the actual effort required. Finally, we emphasize the value of bespoke power simulations as research contributions and their alignment with open science and preregistration practices.

1.  The Need for Bespoke Power Simulations in Human-AI Interaction Research: Human-AI interaction research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or simple helper programs. Bespoke power simulations offer a solution by providing Human-AI interaction researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

@@

Bespoke power simulations play a critical role in human-AI interaction research, allowing researchers to tailor power estimation to the unique aspects of their experiments. The skills required to perform these simulations should be taught to Human-AI interaction researchers, fostering a deeper understanding of statistical power and enhancing research design. It is essential to bridge the gap between perceived and actual effort associated with bespoke power simulations and recognize their value as research contributions. By integrating bespoke power simulations with open science and preregistration practices, researchers can improve the robustness and transparency of their findings, advancing the field.

>>>>

---
title             : "Tutorial for Customized Power Simulations and Data Analyses for Human-AI Interaction Experiments using Generalized Linear Mixed Models"
shorttitle        : "GLMM Power Simulation Tutorial"

@@

affiliation:
  - id            : "1"
    institution   : "LMU Munich, Center for Leadership and People Management"
  - id            : "2"
    institution   : "LMU Munich, Department of Psychology"
  - id            : "3"
    institution   : "University of St. Gallen, Institute of Behavioral Science & Technology"
  - id            : "4"
    institution   : "University College London"

@@

abstract: |
  Understanding how humans interact with articifical intelligence trough experiments becomes increasingly relevant. However, human-AI interaction researchers lack the appropriate tools to conduct power and data analyses for the required complex study designs. In this work, we provide a tutorial on how to run customized power analyses using data simulation based on Generalized Linear Mixed Models (GLMMs). By providing code in a case study, we equip human-AI interaction researchers to simulate their own data and run analyses based on generalized linear mixed models (GLMM). We discuss the outlook.

keywords          : "generalized linear mixed model, data simulation, sample size, power analysis, human-AI interaction"

@@

<!-- General intro, relevance of Human AI Interaction experiments, problem of underpowered & inappropriately analyzed work -->

As the integration of artificial intelligence (AI) into our daily lives continues to increase, it has become increasingly important to study how humans interact with these intelligent systems. This is particularly important as AI becomes more sophisticated and is used in decision-making processes, for instance in medicine, that have significant impacts on individuals and society as a whole. Therefore, thorough experiments are needed to study human-AI interaction and understand how people perceive and respond to these technologies. When conducting research in this field, it is essential to determine an appropriate sample size to ensure that the results obtained are both statistically significant and reliable. Therefore, running sample size estimations is crucial for designing experiments that can provide meaningful insights into human-AI interaction.

@@

<!-- Power simulations are hard! (especially for GLMMs) -->
Power simulations can be challenging, particularly when analyzing complex study designs using (generalized) linear mixed models (GLMMs). While user-friendly software is available for simple statistical models like t-tests, ANOVA, and linear regression, the increasing use of GLMMs in psychological research introduces greater complexity to power calculations. For GLMMs, simulations are often required, necessitating assumptions for both fixed and random effects. However, most tutorials on GLMMs primarily focus on the most common designs, limiting the guidance available for researchers facing more intricate scenarios. Moreover, existing tutorials often rely on heuristics for estimating random effect standard deviations and may incorporate meta-analytic results for both random and fixed effects, further adding to the challenge of conducting accurate power calculations for GLMMs.

@@

<!-- describe the lme4 package and how it works-->
The lme4 R package is a state-of-the-art tool for fitting frequentist Generalized Linear Mixed Models (GLMMs). It provides extensive capabilities for modeling complex data structures, including models with normally distributed random intercepts and random slopes. The package includes a useful function called "simulate" that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling power simulations and other related analyses.

@@

Human-AI interaction research requires meticulous planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and helper programs can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. In this paper, we discuss the necessity of bespoke power simulations in human-AI interaction research, their implications for research design, and the importance of teaching these skills to Human-AI interaction researchers. Furthermore, we highlight the discrepancy between the perceived effort involved in performing bespoke power simulations and the actual effort required. Finally, we emphasize the value of power simulations as research contributions and their alignment with open science and preregistration practices.

1.  The Need for Power Simulations in Human-AI Interaction Research: Human-AI interaction research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or simple helper programs. Bespoke power simulations offer a solution by providing Human-AI interaction researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

@@

Bespoke power simulations play a critical role in human-AI interaction research, allowing researchers to tailor power estimation to the unique aspects of their experiments. The skills required to perform these simulations should be taught to Human-AI interaction researchers, fostering a deeper understanding of statistical power and enhancing research design. It is essential to bridge the gap between perceived and actual effort associated with power simulations and recognize their value as research contributions. By integrating bespoke power simulations with open science and preregistration practices, researchers can improve the robustness and transparency of their findings, advancing the field.
